A comparison of design-based and model-based approaches for finite population spatial data -- Supporting Information.

### 2021-12-22 inst/manuscript/manuscript.tex
more minor updates to incorporate mean standard error addition

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualations and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between
    these approaches is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used. This suggests that
    the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
is IRS-Design. In IRS-Design, samples are selected using the Independent
Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial
locations of the population units, which implies IRS samples are not
spatially balanced. In IRS-Design, samples are analyzed using the
design-based approach with an IRS variance estimator that does not
incorporate the spatial locations of the units in the sample. The second
sampling-analysis combination is IRS-Model, where samples are selected
using the IRS algorithm and analyzed using the model-based approach via
Restricted Maximum Likelihood (REML) estimation (Harville, 1977;
Patterson and Thompson, 1971; Wolfinger et al., 1994). The third
sampling-analysis combination is GRTS-Design, where samples are selected
using the GRTS algorithm and analyzed using the design-based approach
with the local neighborhood variance estimator. The fourth and final
sampling-analysis combination is GRTS-Model, where samples are selected
using the GRTS algorithm and analyzed using the model-based approach via
REML estimation. These sampling-analysis combinations are also provided
in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS
and GRTS samples, equal inclusion probabilities were assumed for all
population units. When IRS assumes equal inclusion probabilities for all
population units, the algorithm is equivalent to ``simple random
sampling.''

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget), respectively, from
Equation\(~\)\ref{equation:expcov}. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance was 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Then we recorded the bias, squared error, and interval coverage for all
sampling-analysis combinations. After all 2000 trials, we summarized the
long-run performance of the combinations by calculating average bias,
rMS(P)E (root-mean-squared error for the design-based approaches and
root-mean-squared-prediction error for the model-based approaches), and
the proportion of times the true mean is contained in its 95\% interval.
The IRS algorithm, IRS variance estimator, GRTS algorithm, and local
neighborhood variance estimator are available in the \texttt{spsurvey}
\textbf{\textsf{R}} package (Dumelle et al., 2021). FPBK is available in
the \texttt{sptotal} \textbf{\textsf{R}} package (Higham et al., 2021).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). We will analyze mercury concentration data collected at
986 lakes as part of the 2012 NLA. Although we can calculate the true
mean mercury concentration values for these 986 lakes, here we will
explore whether or not we can obtain an adequately precise estimate for
the realized mean mercury concentration if we sample only 100 of the 986
lakes. For each of the four familiar sampling-analysis combinations
(IRS-Design, IRS-Model, GRTS-Design, and GRTS-Model), we estimate
(design-based) or predict (model-based) the mean mercury concentration
and construct 95\% confidence (design-based) or 95\% prediction
(model-based) intervals from this sample of 100 lakes, which we compare
to the true mean mercury concentration from all 986 lakes.

@@

The average bias was nearly zero for all four sampling-analysis
combinations in all 36 scenarios, so we omit a more detailed summary of
those results here. Tables for average bias in all 36 simulation
scenarios are provided in the supporting information.

Fig. \ref{fig:figeff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:figeff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So using the GRTS
algorithm or a model-based analysis does not result in much, if any,
loss in efficiency compared to IRS-Design when there is no spatial
covariance. When there is spatial covariance (Fig. \ref{fig:figeff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to perform
best, followed by GRTS-Design, IRS-Model, and finally IRS-Design, though
the difference in relative rMS(P)E among GRTS-Model, GRTS-Design, and
IRS-Model is relatively small. As the strength of spatial covariance
increases, the gap in rMS(P)E between IRS-Design and the other
sampling-analysis combinations widens. Finally we note that when there
is spatial covariance, IRS-Model outperforms IRS-Design by a large
margin, suggesting that the poor design properties of IRS are largely
mitigated by the model-based analysis. These conclusions are similar to
those observed in the grid location layout, so we omit a grid location
layout figure here. Tables for rMS(P)E in all 36 simulation scenarios
are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:figeff}
\end{figure}

We also studied 95\% interval coverage among the sampling-analysis
combinations. The design-based confidence intervals and model-based
prediction intervals were constructed using the normal distribution.
Justification for this comes from the asymptotic normality of means via
the Central Limit Theorem. Fig. \ref{fig:figconf} shows the 95\%
interval coverage for each of the four sampling-analysis combinations in
the random location layout. Within each scenario, the sampling-analysis
combinations tend to have fairly similar interval coverage. Coverage in
the normal response scenarios was usually near 95\%, while coverage in
the lognormal response scenarios varied from from 90\% to 95\% but
increased with the sample size. At a sample size of 200, all four
sampling-analysis combinations had approximately 95\% interval coverage
in both response scenarios for all dependent error proportions. These
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for interval
coverage in all 36 simulation scenarios are provided in the supporting
information.

@@

Fig. \ref{fig:merc} shows a map and histogram of mercury concentration
in all 986 NLA lakes. The map shows mercury concentration exhibits some
spatial patterning, with high mercury concentrations in the northeast
and north central United States. The histogram shows that mercury
concentration is right-skewed, with most lakes having a low value of
mercury concentration but a few having a much higher concentration. Fig.
\ref{fig:merc} also shows mercury concentration's empirical
semivariogram. The empirical semivariogram can be used as a tool to
visualize spatial dependence. It quantifies the halved squared
differences (semivariance) among mercury concentration at different
distances apart. When a process has spatial covariance (exhibits spatial
dependence), the semivariance tends to be smaller at small distances and
larger at large distances. The empirical semivariogram in Fig.
\ref{fig:merc} suggests that mercury concentration is exhibits spatial
dependence. Lastly we note that the realized mean mercury concentration
in the 986 NLA lakes is 103.2 ng / g.

\begin{figure}
\centering
\begin{subfigure}{0.98\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption*{}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption*{}
  \label{fig:mercury_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/sv_plot.jpeg}
  \caption*{}
  \label{fig:sv_plot}
\end{subfigure}
\caption{Mercury concentration visualizations for the population (Hg) for 986 lakes in the NLA data. A spatial layout is in the top row, a histogram is in the bottom row and left column, and an empirical semivariogram is in the bottom row and right column.}
\label{fig:merc}
\end{figure}

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. For all four sampling-analysis combinations, the true
realized mean mercury concentration is within the bounds of the 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Though we should not generalize these results to other samples from
these data, we do note a couple of patterns. The design-based IRS
analysis shows the largest standard error: a likely reason is that this
is the only approach that does not incorporate any spatial locations.
Additionally, both analyses using GRTS sampling have lower standard
errors than both analyses using IRS sampling.

\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Approach & Est/Pred & SE & 95\% LB & 95\% UB \\ 
  \hline
IRS-Design & 112.7 & 8.8 & 95.4 & 129.9 \\ 
  IRS-Model & 110.5 & 7.9 & 95.0 & 125.9 \\ 
  GRTS-Design & 101.8 & 6.1 & 89.8 & 113.7 \\ 
  GRTS-Model & 102.3 & 5.9 & 90.8 & 113.9 \\ 
   \hline
\end{tabular}
\caption{\label{tab:appliedtab} For each sampling-analysis combination (Approach), estimates/predictions (Est/Pred), standard errors (SE), lower 95\% interval bounds (95\% LB), and upper 95\% interval bounds (95\% UB) for mean mercury concentration computed using a sample of 100 lakes in the NLA data. The true mean concentration of all 986 lakes in the NLA data is 103.2 ng / g.} 
\end{table}

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach
incorporates randomness through sampling to estimate population
parameters. The model-based approach incorporates randomness through
distributional assumptions to predict realized values of a stochastic
process. Though these approaches have often been compared in the
literature from theoretical and analytical perspectives, our
contribution lies in studying them in a spatial context while
implementing spatially balanced sampling and the local neighborhood
variance estimator (in the design-based approach). Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, the sampling decision
(IRS vs GRTS) is most important when using a design-based analysis.
Though GRTS-Model still outperformed IRS-Model, the model-based analysis
mitigated most of the inefficiencies that result from the IRS samples
lacking spatial balance. Second, independent of the analysis approach,
we found no reason to prefer IRS over GRTS for sampling spatial data --
GRTS-Design and GRTS-Model generally performed at least as well as their
IRS counterparts when there was no spatial covariance and noticeably
better than their IRS counterparts when there was spatial covariance.
Third, as the strength of spatial covariance increases, the gap in
rMS(P)E between IRS-Design and the other sampling-analysis combinations
also increases. Fourth and finally, when the response was normal,
interval coverage for all sampling-analysis combinations was very close
to 95\% for all sample sizes; when the response was lognormal, interval
coverage for all sampling and analysis was between 90\% and 95\% and
closest to 95\% when \(n = 200\).

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference lie on fundamentally different foundations. In the
    design-based approach, inference depends on random sampling. In the
    model-based approach, inference depends on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulations and an
    analysis of real mercury concentration data. In the simulations, a
    variety of sample sizes, location layouts, dependence structures, and
    response types are considered. In the simualations and real data
    analysis, the population mean is the parameter of interest and
    performance is measured using statistics like bias, squared error, and
    interval length and coverage.
  \item
    When studying the simulations and mercury concentration data, we found
    that regardless of the strength of spatial dependence in the data,
    sampling plans that incorporate spatial locations (spatially balanced
    samples) generally outperform sampling plans that ignore spatial
    locations (non-spatially balanced samples). We also found that
    model-based approaches tend to outperform design-based approaches,
    even when the data are skewed (and by consequence, the model-based
    distributional assumptions violated). The performance gap between
    these approaches is small when spatially balanced samples are used but
    large when non-spatially balanced samples are used. This suggests that
    the sampling choice (whether to select a sample that is spatially
    balanced) is most important when performing design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial data that
    practitioners must consider when choosing between them. We provide
    relevant background contextualizing each approach and study their
    properties in a variety of scenarios, making recommendations for use
    based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

We used a simulation study to investigate performance of four
sampling-analysis combinations. The first sampling-analysis combination
is IRS-Design. In IRS-Design, samples are selected with the Independent
Random Sampling (IRS) algorithm. The IRS algorithm ignores the spatial
locations of the population units, thus IRS samples are not spatially
balanced. In IRS-Design, samples are analyzed using the design-based
approach via the Horvitz-Thompson mean estimator and an IRS variance
estimator that does not incorporate the spatial locations of the units
in the sample. The second sampling-analysis combination is IRS-Model,
where samples are selected with the IRS algorithm and analyzed using the
model-based approach via Restricted Maximum Likelihood (REML) estimation
(Harville, 1977; Patterson and Thompson, 1971; Wolfinger et al., 1994).
The third sampling-analysis combination is GRTS-Design, where samples
are selected with the GRTS algorithm and analyzed using the design-based
approach via the Horvitz-Thompson mean estimator and the local
neighborhood variance estimator (which does incorporate the spatial
locations of the units in the sample). The fourth and final
sampling-analysis combination is GRTS-Model, where samples are selected
with the GRTS algorithm and analyzed using the model-based approach via
REML estimation. These sampling-analysis combinations are also provided
in Table \ref{tab:designanalysis}. Lastly we note that for both the IRS
and GRTS samples, equal inclusion probabilities were assumed for all
population units. When IRS assumes equal inclusion probabilities for all
population units, the algorithm is equivalent to simple random sampling
(SRS).

@@

Performance for the four sampling-analysis combinations was evaluated in
36 different simulation scenarios. The 36 scenarios resulted from the
crossing of three sample sizes, two location layouts (of the population
units), two response types, and three proportions of dependent random
error. The three sample sizes (\(n\)) were \(n = 50, n = 100,\) and
\(n = 200\). Samples were always selected from a population size (\(N\))
of \(N = 900\). The two location layouts were random and gridded.
Locations in the random layout were randomly generated inside the unit
square (\([0, 1] \times [0, 1]\)). Locations in the gridded layout were
placed on a fixed, equally spaced grid inside the unit square. The two
response types were normal and lognormal. For the normal response type,
the response was simulated using mean-zero random errors with the
exponential covariance (Equation\(~\)\ref{equation:expcov}) for varying
proportions of dependent random error. The proportion of dependent
random error is represented by
\(\sigma^2_1 / (\sigma^2_1 + \sigma^2_2)\), where \(\sigma^2_1\) and
\(\sigma^2_2\) are the dependent random error variance (partial sill)
and independent random error variance (nugget) from
Equation\(~\)\ref{equation:expcov}, respectively. The total variance,
\(\sigma^2_1 + \sigma^2_2\), was always 2. The range was always
\(\sqrt{2} / 3\), which means that the correlation in the dependent
random error decayed to nearly zero at the largest possible distance
between two population units in the domain. For the lognormal response
type, the response was first simulated using the same approach as for
the normal response type, except that the total variance was 0.6931
instead of 2. The response was then exponentiated, yielding a lognormal
random variable whose total variance was 2. The lognormal responses were
used to evaluate performance of the sampling-analysis approaches for
data that were skewed (i.e., not normal).

@@

In each of the 36 simulation scenarios, there were 2000 independent
simulation trials. In each trial, IRS and GRTS samples were selected and
then design-based and model-based analyses were used to estimate
(design-based) or predict (model-based) the mean and construct 95\%
confidence (design-based) or 95\% prediction (model-based) intervals.
Then we recorded the bias, squared error, standard error, and interval
coverage for all sampling-analysis combinations. After all 2000 trials,
we summarized the long-run performance of the combinations by
calculating mean bias, rMS(P)E (root-mean-squared error for the
design-based approaches and root-mean-squared-prediction error for the
model-based approaches), MStdE (mean standard error), and the proportion
of times the true mean is contained in its 95\% confidence
(design-based) or 95\% prediction (model-based) interval. The 95\%
intervals were constructed using the normal distribution. Justification
for this comes from the asymptotic normality of means via the Central
Limit Theorem (under some assumptions). Quantifying mean bias and
rMS(P)E is important because they help us understand how far (under
different loss metrics) the estimates (design-based) or predictions
(model-based) tend to be from the true mean. Quantifying MStdE is
important because it helps us understand how precise intervals tend to
be. Quantifying interval coverage is important because it helps us
understand how often our 95\% intervals actually contain the true mean.

The IRS algorithm, IRS variance estimator, GRTS algorithm, and local
neighborhood variance estimator are available in the \texttt{spsurvey}
\textbf{\textsf{R}} package (Dumelle et al., 2021). FPBK is available in
the \texttt{sptotal} \textbf{\textsf{R}} package (Higham et al., 2021).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality
(USEPA, 2012). We will analyze mercury concentration data collected at
986 lakes from the 2012 NLA. Although we can calculate the true mean
mercury concentration values for these 986 lakes, here we will explore
whether or not we can obtain an adequately precise estimate
(design-based) or prediction (model-based) for the realized mean mercury
concentration if we sample only 100 of the 986 lakes. For each of the
four familiar sampling-analysis combinations (IRS-Design, IRS-Model,
GRTS-Design, and GRTS-Model), we estimate (design-based) or predict
(model-based) the mean mercury concentration and construct 95\%
intervals from this sample of 100 lakes and compare to the true mean
mercury concentration from all 986 lakes.

@@

The mean bias was nearly zero for all four sampling-analysis
combinations in all 36 scenarios, so we omit a more detailed summary of
those results here. Tables for mean bias in all 36 simulation scenarios
are provided in the supporting information.

Fig. \ref{fig:rmspe_eff} shows the relative rMS(P)E of the four sampling
analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative rMS(P)E is defined as
\begin{equation*}
\frac{\text{rMS(P)E of sampling-analysis combination}}{\text{rMS(P)E of IRS-Design}},
\end{equation*} When there is no spatial covariance (Fig.
\ref{fig:rmspe_eff}, ``Prop DE: 0'' row), the four sampling-analysis
combinations have approximately equal rMS(P)E. So when there is no
spatial covariance, using the GRTS algorithm or a model-based analysis
does not result in much, if any, loss in efficiency compared to
IRS-Design. When there is spatial covariance (Fig. \ref{fig:rmspe_eff},
``Prop DE: 0.5'' and ``Prop DE: 0.9'' rows), GRTS-Model tends to have
the lowest rMS(P)E, followed by GRTS-Design, IRS-Model, and finally
IRS-Design, though the difference in relative rMS(P)E among GRTS-Model,
GRTS-Design, and IRS-Model is relatively small. As the strength of
spatial covariance increases, the gap in rMS(P)E between IRS-Design and
the other sampling-analysis combinations widens. Finally we note that
when there is spatial covariance, IRS-Model has a much lower rMS(P)E
than IRS-Design, suggesting that the poor design properties of IRS are
largely mitigated by the model-based analysis. These rMS(P)E conclusions
are similar to those observed in the grid location layout, so we omit a
grid location layout figure here. Tables for rMS(P)E in all 36
simulation scenarios are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative rMS(P)E in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:rmspe_eff}
\end{figure}

Fig. \ref{fig:mse_eff} shows the relative MStdE of the four
sampling-analysis combinations using the random location layout with
``IRS-Design'' as the baseline. The relative MStdE is defined as
\begin{equation*}
\frac{\text{MStdE of sampling-analysis combination}}{\text{MStdE of IRS-Design}},
\end{equation*} Many general takeaways regarding MStdE are similar to
general takeaways regarding rMS(P)E: there seems to be no benefit to
using IRS, even when there is no spatial covariance; as the strength of
spatial covariance increases, the gap in MStdE between IRS-Design and
the other sampling-analysis combinations widens; and IRS-Model
outperforms (has a much lower MStdE) IRS-Design by a large margin. These
fact that the rMS(P)E and MStdE findings are similar is not particularly
surprising because the mean bias for all sampling-analysis combinations
was nearly zero, thus rMS(P)E is driven by the standard error of the
estimators (design-based) or predictors (model-based). We do note that
between GRTS-Design and GRTS-Model, GRTS-Design had lower MStdE when
there was no spatial covariance or a medium amount of spatial covariance
(Fig. \ref{fig:mse_eff}, ``Prop DE: 0'' and ``Prop DE: 0.5'' rows) and
GRTS-Model had lower MStdE when there was a high amount of spatial
covariance (Fig. \ref{fig:mse_eff}, ``Prop DE: 0.9'' row). These MStdE
conclusions are similar to those observed in the grid location layout,
so we omit a grid location layout figure here. Tables for MStdE in all
36 simulation scenarios are provided in the supporting information.

\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mse_eff.jpeg}
  \caption{Relative MStdE in the simulation study for the four sampling-analysis combinations. The rows indicate the proportion of dependent error and the columns indicate the response type.}
  \label{fig:mse_eff}
\end{figure}

Fig. \ref{fig:figconf} shows the 95\% interval coverage for each of the
four sampling-analysis combinations in the random location layout.
Within each scenario, the sampling-analysis combinations tend to have
fairly similar interval coverage, though when \(n = 50\) or \(n = 100\),
GRTS-Design coverage is usually a few percentage points lower than the
other combinations. Coverage in the normal response scenarios was
usually near 95\%, while coverage in the lognormal response scenarios
usually varied from 90\% to 95\% but increased with the sample size. At
a sample size of 200, all four sampling-analysis combinations had
approximately 95\% interval coverage in both response scenarios for all
dependent error proportions. These interval coverage conclusions are
similar to those observed in the grid location layout, so we omit a grid
location layout figure here. Tables for interval coverage in all 36
simulation scenarios are provided in the supporting information.

@@

Fig. \ref{fig:merc} shows a map and histogram of mercury concentration
in all 986 NLA lakes. The map shows mercury concentration exhibits some
spatial patterning, with high mercury concentrations in the northeast
and north central United States. The histogram shows that mercury
concentration is right-skewed, with most lakes having a low value of
mercury concentration but a few having a much higher concentration. Fig.
\ref{fig:merc} also shows mercury concentration's empirical
semivariogram. The empirical semivariogram can be used as a tool to
visualize spatial dependence. It quantifies the mean of the halved
squared differences (semivariance) among all pairs of mercury
concentrations at different distances apart. When a process has spatial
covariance (exhibits spatial dependence), the mean semivariance tends to
be smaller at small distances and larger at large distances. The
empirical semivariogram in Fig. \ref{fig:merc} suggests that mercury
concentration exhibits spatial dependence. Lastly we note that the true
mean mercury concentration in the 986 NLA lakes is 103.2 ng / g.

\begin{figure}
\centering
\begin{subfigure}{0.98\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_map.jpeg}
  \caption*{}
  \label{fig:mercury_map}
\end{subfigure} \\
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/mercury_hist.jpeg}
  \caption*{}
  \label{fig:mercury_hist}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/sv_plot.jpeg}
  \caption*{}
  \label{fig:sv_plot}
\end{subfigure}
\caption{Mercury concentration (Hg) visualizations for all 986 lakes in the NLA data. A spatial layout is in the top row, a histogram is in the bottom row and left column, and an empirical semivariogram is in the bottom row and right column.}
\label{fig:merc}
\end{figure}

We selected a single IRS sample and a single GRTS sample and estimated
(design-based) or predicted (model-based) the mean mercury concentration
and constructed 95\% confidence (design-based) and 95\% (model-based)
prediction intervals. For the model-based analyses, the exponential
covariance was used. Table \ref{tab:appliedtab} shows the results from
these analyses. Though we should not generalize these results to other
samples from these data, we do mention a few findings. First, IRS-Design
has the largest standard error. Second, compared to IRS-Design and
IRS-Model, GRTS-Design and GRTS-Model are much closer to the true mean
mercury concentration (have bias closer to zero) and have much lower
standard errors (more precise intervals). Third, GRTS-Model has the
least amount of bias and the lowest standard error (most precise
interval). Finally, we note that for all sampling-analysis combinations,
the true mean mercury concentration (103.2 ng / g) is within the bounds
of the combination's 95\% interval.

\begin{table}[ht]
\centering
\begin{tabular}{lrrrrr}
  \hline
Approach & True Mean & Est/Pred & SE & 95\% LB & 95\% UB \\ 
  \hline
IRS-Design & 103.2 & 112.7 & 8.8 & 95.4 & 129.9 \\ 
  IRS-Model & 103.2 & 110.5 & 7.9 & 95.0 & 125.9 \\ 
  GRTS-Design & 103.2 & 101.8 & 6.1 & 89.8 & 113.7 \\ 
  GRTS-Model & 103.2 & 102.3 & 5.9 & 90.8 & 113.9 \\ 
   \hline
\end{tabular}
\caption{\label{tab:appliedtab} For each sampling-analysis combination (Approach), the true mean mercury concentration (True Mean), estimates/predictions (Est/Pred), standard errors (SE), lower 95\% interval bounds (95\% LB), and upper 95\% interval bounds (95\% UB) for mean mercury concentration computed using a sample of 100 lakes in the NLA data.} 
\end{table}

@@

The design-based and model-based approaches to statistical inference are
fundamentally different paradigms. The design-based approach relies on
random sampling to estimate population parameters. The model-based
approach relies on distributional assumptions to predict realized values
of a stochastic process. Though the model-based approach does not rely
on random sampling, it can still be beneficial as a way to guard against
preferential sampling. The design-based and model-based approaches have
often been compared in the literature from theoretical and analytical
perspectives, but our contribution lies in studying them in a spatial
context while implementing spatially balanced sampling and the
design-based, local neighborhood variance estimator. Aside from the
theoretical differences described, a few analytical findings from the
simulation study are particularly notable. First, independent of the
analysis approach, we found no reason to prefer IRS over GRTS when
sampling spatial data -- GRTS-Design and GRTS-Model generally had
similar rMS(P)E as their IRS counterparts when there was no spatial
covariance and lower rMS(P)E than their IRS counterparts when there was
spatial covariance. Second, the sampling decision (IRS vs GRTS) is most
important when using a design-based analysis. Though GRTS-Model still
had lower rMS(P)E than IRS-Model, the model-based analysis mitigated
most of the rMS(P)E inefficiencies that result from the IRS samples
lacking spatial balance. Third, as the strength of spatial covariance
increases, the gap in rMS(P)E and MStdE between IRS-Design and the other
sampling-analysis combinations also increases, likely because IRS-Design
is the only combination that ignores spatial locations in sampling and
analysis. Fourth and finally, when the response was normal, interval
coverage for all sampling-analysis combinations was usually close to
95\% for all sample sizes; when the response was lognormal, interval
coverage for all sampling-analysis combinations was usually between 90\%
and 95\% and closest to 95\% when \(n = 200\).

### 2021-12-22 inst/manuscript/supporting.Rmd
more minor updates to incorporate mean standard error addition

Next we present tables summarizing mean estimation bias (design-based) or mean prediction bias (model-based), root-mean-squared error (design-based) or root-mean-squared-prediction error (model-based), and 95% confidence interval coverage (design-based) or 95% prediction interval coverage (model-based) for all 36 simulation scenarios.
```{r}
make_sim_table <- function(table_value) {
  sim_table <- read_csv(here("inst", "output", "simulation_summary", paste0("allsims", table_value, ".csv")))
  sim_table <- sim_table %>%
    dplyr::select(Approach = approach,
                  Layout = gridded,
                  Response = resptype,
                  `DE Prop` = psill_ratio,
                  n = n,
                  Bias = mpbias,
                  `rMS(P)E` = rmspe,
                  Coverage = coverage) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-Design",
       Approach == "Design IRS" ~ "IRS-Design",
       Approach == "Model GRTS" ~ "GRTS-Model",
       Approach == "Model IRS" ~ "IRS-Model"
     ),
     Layout = case_when(
        Layout ~ "Gridded",
        !Layout ~ "Random"
      ),
     Response = case_when(
       Response == "normal" ~ "Normal",
       Response == "lognormal" ~ "Lognormal"
     ),
     across(c(`DE Prop`, n), as.character)
    )
sim_table <- sim_table[c(2, 4, 1, 3), ]
sim_table_out <- xtable(
  sim_table,
  digits = 4,
  caption = paste0("Sampling-analysis combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (DE Prop), sample size (n), mean (prediction) bias (Bias), root-mean-squared-(prediction) error (rMS(P)E), and 95\\% interval coverage (Coverage) in simulation scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("llllrrrrr")
)
print(sim_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.85")
}
```

>>>>

Next we present tables summarizing mean estimation bias (design-based) or mean prediction bias (model-based), mean standard errors, root-mean-squared error (design-based) or root-mean-squared-prediction error (model-based), and 95% confidence interval coverage (design-based) or 95% prediction interval coverage (model-based) for all 36 simulation scenarios.
```{r}
make_sim_table <- function(table_value) {
  sim_table <- read_csv(here("inst", "output", "simulation_summary", paste0("allsims", table_value, ".csv")))
  sim_table <- sim_table %>%
    dplyr::select(Approach = approach,
                  Layout = gridded,
                  Response = resptype,
                  PDE = psill_ratio,
                  n = n,
                  Bias = mpbias,
                  `rMS(P)E` = rmspe,
                  MStdE = mse,
                  Coverage = coverage
                  ) %>%
    mutate(
     Approach = case_when(
       Approach == "Design GRTS" ~ "GRTS-Design",
       Approach == "Design IRS" ~ "IRS-Design",
       Approach == "Model GRTS" ~ "GRTS-Model",
       Approach == "Model IRS" ~ "IRS-Model"
     ),
     Layout = case_when(
        Layout ~ "Gridded",
        !Layout ~ "Random"
      ),
     Response = case_when(
       Response == "normal" ~ "Normal",
       Response == "lognormal" ~ "Lognormal"
     ),
     across(c(PDE, n), as.character)
    )
sim_table <- sim_table[c(2, 4, 1, 3), ]
sim_table_out <- xtable(
  sim_table,
  digits = 4,
  caption = paste0("Sampling-analysis combination (Approach), population layout (Layout), response type (Response), proportion of dependent random error (PDE), sample size (n), mean (prediction) bias (Bias), root-mean-squared-(prediction) error (rMS(P)E), mean standard error (MStdE), and 95\\% interval coverage (Coverage) in simulation scenario ", table_value, "."),
   type = "latex",
  latex.environments = "center",
  align = c("llllrrrrrr")
)
print(sim_table_out, include.rownames = FALSE, comment = FALSE, scalebox = "0.8")
}
```
